Component response rate variation drives stability in large complex systems

### 2018-07-30 notebook/ms.tex
Rework all files for arXiv

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{setspace}
\setstretch{1}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\hypersetup{unicode=true,
            pdftitle={Component response rate variation drives stability in large complex systems},
            pdfauthor={A. Bradley Duthie},
            colorlinks=true,
            linkcolor=blue,
            citecolor=Blue,
            urlcolor=Blue,
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Component response rate variation drives stability in large complex
systems}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{A. Bradley Duthie}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{Biological and Environmental Sciences, University of Stirling, Stirling,
UK, FK9 4LA
\href{mailto:alexander.duthie@stir.ac.uk}{\nolinkurl{alexander.duthie@stir.ac.uk}}}

\usepackage{amsmath}
\usepackage{natbib}
\usepackage{lineno}
\usepackage[utf8]{inputenc}
\linenumbers
\bibliographystyle{amnatnat}

@@

@@

\textbf{The stability of a complex system generally decreases with
increasing system size, as is demonstrated by random matrix
theory\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}.
This counter-intuitive result, first shown by
May\textsuperscript{\protect\hyperlink{ref-May1972}{1}}, is broadly
relevant for understanding the dynamics and persistence of systems such
as
ecological\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}},
neurological\textsuperscript{\protect\hyperlink{ref-Gray2008}{3},\protect\hyperlink{ref-Gray2009}{4}},
biochemical\textsuperscript{\protect\hyperlink{ref-Rosenfeld2009}{5},\protect\hyperlink{ref-MacArthur2010}{6}}
and
socio-economic\textsuperscript{\protect\hyperlink{ref-Haldane2011}{7}--\protect\hyperlink{ref-Bardoscia2017}{9}}
networks. Much attention has especially been given to the stability of
ecological communities such as food webs or mutualist networks, with
recent work investigating how different community structures affect
stability\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Mougi2012}{10}--\protect\hyperlink{ref-Patel2018}{14}}.
But more broadly, stabilising mechanisms in complex systems remain
under-developed, and the effect of variation in the response rate of
individual system components remains an open
problem\textsuperscript{\protect\hyperlink{ref-Allesina2015}{15}}. Here
I show that when components of a complex system respond to system
dynamics at different rates (\(\boldsymbol{\gamma}\)), the potential for
system stability is markedly increased. Stability is caused by the
clustering of some eigenvalues toward the centre of eigenvalue
distributions despite the destabilising effect of higher interaction
strength variation (\(\boldsymbol{\sigma^{2}}\)). This effect of
variation in \(\boldsymbol{\gamma}\) becomes increasingly important as
system size increases, to the extent that the largest stable complex
systems would otherwise be unstable if not for
\(\boldsymbol{Var(\gamma)}\). My results therefore reveal a previously
unconsidered driver of system stability that is likely to be pervasive
across all complex systems. Future research in complex systems should
therefore account for the varying response rates of individual system
components when assessing whole system stability.}

In 1972, May\textsuperscript{\protect\hyperlink{ref-May1972}{1}} first
demonstrated that randomly assembled systems of sufficient complexity
are almost inevitably unstable given infinitesimally small
perturbations. Complexity in this case is defined by the size of the
system (i.e., the number of potentially interacting components; \(S\)),
its connectance (i.e., the probability that one component will interact
with another; \(C\)), and the variance of interaction strengths
(\(\sigma^{2}\))\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}.
May's finding that the probability of local stability falls to near zero
given a sufficiently high threshold of \(\sigma\sqrt{SC}\) has profound
consequences across multiple disciplines, raising the question of how
complex systems in, e.g.,
ecology\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Mougi2012}{10},\protect\hyperlink{ref-Grilli2017}{13},\protect\hyperlink{ref-Allesina2015}{15}}
or
banking\textsuperscript{\protect\hyperlink{ref-Haldane2011}{7},\protect\hyperlink{ref-Bardoscia2017}{9},\protect\hyperlink{ref-May2008}{16}}
are predicted to persist or change.

Randomly assembled complex systems can be represented as large square
matrices (\(M\)) with \(S\) components (e.g.,
species\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}} or
banks\textsuperscript{\protect\hyperlink{ref-Haldane2011}{7}}). One
element of such a matrix, \(M_{ij}\), defines how component \(j\)
affects component \(i\) in the system at a point of
equilibrium\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}.
Off-diagonal elements (\(i \neq j\)) therefore define interactions
between components, while diagonal elements (\(i = j\)) define component
self-regulation (e.g., carrying capacity in ecological communities).
Traditionally, values of off-diagonal elements are assigned non-zero
values with a probability \(C\), which are sampled from a distribution
with variance \(\sigma^{2}\); diagonal elements are set to
-1\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{15}}.
Local system stability is assessed using eigenanalysis, with the system
being stable if the real parts of all eigenvalues (\(\lambda\)) of \(M\)
are negative
(\(\max\left(\Re(\lambda)\right) < 0\))\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}.
In a large system (high \(S\)), eigenvalues are distributed
uniformly\textsuperscript{\protect\hyperlink{ref-Tao2010}{17}} within a
circle centred at \(\Re = -1\) (the mean value of diagonal elements) and
\(\Im = 0\), with a radius of
\(\sigma\sqrt{SC}\)\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{15}}
(Figs 1a and 2a). Local stability of randomly assembled systems
therefore becomes increasingly unlikely as \(S\), \(C\), and
\(\sigma^{2}\) increase.

The above stability criterion assumes that individual components respond
to perturbations of the system at the same rate (\(\gamma\)), but this
is highly unlikely in any complex system. In ecological communities, for
example, the rate at which population density changes following
perturbation will depend on the generation time of individuals, which
might vary by orders of magnitude among species. Species with short
generation times will respond quickly (high \(\gamma\)) to perturbations
relative to species with long generation times (low \(\gamma\)).
Similarly, the speed at which individual banks respond to perturbations
in financial networks, or individuals or institutions respond to
perturbations in complex social networks, is likely to vary. The effect
of such variance has not been investigated in complex systems theory.
Intuitively, variation in \(\gamma\) might be expected to decrease
system stability by introducing a new source of variation into the
system and thereby increasing \(\sigma\). Here I show why, despite
higher \(\sigma\), complex systems in which \(\gamma\) varies are
actually more likely to be stable, especially when \(S\) is high.

@@

\textbf{Figure 1: Example distribution of eigenvalues before (a) and
after (b) separating a randomly generated complex system into fast
(\(\boldsymbol{\gamma} = 1.95\)) and slow
(\(\boldsymbol{\gamma} = 0.05\)) component response rates.} Each panel
shows the same system where \(S = 200\), \(C = 0.05\), and
\(\sigma = 0.4\), and in each case \(E[\gamma] = 1\) (i.e., only the
distribution of \(\gamma\) differs between panels). \textbf{a.}
Eigenvalues plotted when all \(\gamma = 1\); distributions of points are
uniformly distributed within the grey circle with a radius of
\(\sigma\sqrt{SC} =\) 1.238 centred at -1 on the real axis. \textbf{b.}
Eigenvalues plotted when half \(\gamma = 1.95\) and half
\(\gamma = 0.05\); distributions of points can be partitioned into one
large circle of \(\sigma\sqrt{SC} =\) 1.718 centred at
\(\gamma = -1.95\) and one small circle of \(\sigma\sqrt{SC} =\) 0.044
centred at \(\gamma = -0.05\). In a, the maximum real eigenvalue
\(\max\left(\Re(\lambda)\right) =\) 0.2344871, while in b
\(\max\left(\Re(\lambda)\right) =\) -0.0002273135, meaning that the
complex system in b but not a is stable because in b
\(\max\left(\Re(\lambda)\right) < 0\). In 1 million randomly generated
complex systems under the same parameter values, 1 was stable when
\(\gamma = 1\) while 32 were stable when \(\gamma = \{1.95, 0.05\}\).
Overall, complex systems that are separated into fast versus slow
components tend to be more stable than otherwise identical systems with
identical component response rates.

\includegraphics{ms_files/figure-latex/unnamed-chunk-4-1.pdf}

@@

\textbf{Figure 2: Distributions of eigenvalues before (a) and after (b)
introducing variation in component response rate
(\(\boldsymbol{\gamma}\)) in complex systems.} Each panel show the same
system where \(S = 1000\), \(C = 1\), and \(\sigma = 0.4\). \textbf{a.}
Eigenvalues plotted in the absence of \(Var(\gamma)\) where
\(E[\gamma] = 1\), versus \textbf{b.} eigenvalues plotted given
\(\gamma \sim \mathcal{U}(0, 2)\), which increases the variance of
interaction strengths (\(\sigma^{2}\)) but clusters eigenvalues toward
the distribution's centre (-1, 0). Black and red elipses in both panels
show the circle centred on the distribution in panels a and b,
respectively, which have a radius of \(\sigma \sqrt{SC}\). Proportions
of \(\Re(\lambda) < 0\) are 0.549 and 0.547 for a and b, respectively.

\includegraphics{ms_files/figure-latex/unnamed-chunk-7-1.pdf}

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{setspace}
\setstretch{1}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\hypersetup{unicode=true,
            pdftitle={Component response rate variation drives stability in large complex systems},
            pdfauthor={A. Bradley Duthie},
            colorlinks=true,
            linkcolor=blue,
            citecolor=Blue,
            urlcolor=Blue,
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\ImportTok}[1]{{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{{#1}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{{#1}}}}
\newcommand{\BuiltInTok}[1]{{#1}}
\newcommand{\ExtensionTok}[1]{{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{{#1}}}}
\newcommand{\NormalTok}[1]{{#1}}
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Component response rate variation drives stability in large complex
systems}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{A. Bradley Duthie (
\href{mailto:alexander.duthie@stir.ac.uk}{\nolinkurl{alexander.duthie@stir.ac.uk}}
)}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{Biological and Environmental Sciences, University of Stirling, Stirling,
UK, FK9 4LA}

\usepackage{amsmath}
\usepackage{natbib}
\usepackage[utf8]{inputenc}
\bibliographystyle{amnatnat}

@@

\textbf{The stability of a complex system generally decreases with
increasing system size and interconnectivity, a counterintuitive result
of widespread importance across the physical, life, and social sciences.
Despite recent interest in the relationship between system properties
and stability, the effect of variation in the response rate of
individual system components remains unconsidered. Here I vary the
component response rates (\(\boldsymbol{\gamma}\)) of randomly generated
complex systems. I show that when component response rates vary, the
potential for system stability is markedly increased. Variation in
\(\boldsymbol{\gamma}\) becomes increasingly important as system size
increases, such that the largest stable complex systems would be
unstable if not for \(\boldsymbol{Var(\gamma)}\). My results reveal a
previously unconsidered driver of system stability that is likely to be
pervasive across all complex systems.}

In 1972, May\textsuperscript{\protect\hyperlink{ref-May1972}{1}} first
demonstrated that randomly assembled systems of sufficient complexity
are almost inevitably unstable given infinitesimally small
perturbations. Complexity in this case is defined by the size of the
system (i.e., the number of potentially interacting components; \(S\)),
its connectance (i.e., the probability that one component will interact
with another; \(C\)), and the variance of interaction strengths
(\(\sigma^{2}\))\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}.
May's finding that the probability of local stability falls to near zero
given a sufficiently high threshold of \(\sigma\sqrt{SC}\) is broadly
relevant for understanding the dynamics and persistence of systems such
as
ecological\textsuperscript{\protect\hyperlink{ref-May1972}{1}--\protect\hyperlink{ref-Grilli2017}{5}},
neurological\textsuperscript{\protect\hyperlink{ref-Gray2008}{6},\protect\hyperlink{ref-Gray2009}{7}},
biochemical\textsuperscript{\protect\hyperlink{ref-Rosenfeld2009}{8},\protect\hyperlink{ref-MacArthur2010}{9}},
and
socio-economic\textsuperscript{\protect\hyperlink{ref-May2008}{10}--\protect\hyperlink{ref-Bardoscia2017}{13}}
networks. As such, identifying general principles that drive stability
in complex systems is of wide-ranging importance.

Randomly assembled complex systems can be represented as large square
matrices (\(\mathbf{M}\)) with \(S\) components (e.g., networks of
species\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}} or
banks\textsuperscript{\protect\hyperlink{ref-Haldane2011}{11}}). One
element of such a matrix, \(M_{ij}\), defines how component \(j\)
affects component \(i\) in the system at a point of
equilibrium\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}.
Off-diagonal elements (\(i \neq j\)) therefore define interactions
between components, while diagonal elements (\(i = j\)) define component
self-regulation (e.g., carrying capacity in ecological communities).
Traditionally, off-diagonal elements are assigned non-zero values with a
probability \(C\), which are sampled from a distribution with variance
\(\sigma^{2}\); diagonal elements are set to
-1\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{4}}.
Local system stability is assessed using eigenanalysis, with the system
being stable if the real parts of all eigenvalues (\(\lambda\)) of
\(\mathbf{M}\) are negative
(\(\max\left(\Re(\lambda)\right) < 0\))\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}.
In a large system (high \(S\)), eigenvalues are distributed
uniformly\textsuperscript{\protect\hyperlink{ref-Tao2010}{14}} within a
circle centred at \(\Re = -1\) (the mean value of diagonal elements) and
\(\Im = 0\), with a radius of
\(\sigma\sqrt{SC}\)\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{4}}
(Figs 1a and 2a). Local stability of randomly assembled systems
therefore becomes increasingly unlikely as \(S\), \(C\), and
\(\sigma^{2}\) increase.

May's\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}
stability criterion \(\sigma\sqrt{SC} < 1\) assumes that individual
components respond to perturbations of the system at the same rate
(\(\gamma\)), but this is highly unlikely in any complex system. In
ecological communities, for example, the rate at which population
density changes following perturbation will depend on the generation
time of organisms, which might vary by orders of magnitude among
species. Species with short generation times will respond quickly (high
\(\gamma\)) to perturbations relative to species with long generation
times (low \(\gamma\)). Similarly, the speed at which individual banks
respond to perturbations in financial networks, or individuals or
institutions respond to perturbations in complex social networks, is
likely to vary. The effect of such variance on stability has not been
investigated in complex systems theory. Intuitively, variation in
\(\gamma\) might be expected to decrease system stability by introducing
a new source of variation into the system and thereby increasing
\(\sigma\). Here I show why, despite higher \(\sigma\), complex systems
in which \(\gamma\) varies are actually more likely to be stable,
especially when \(S\) is high.

\subsection{Results}\label{results}

\textbf{Component response rates of random complex systems}. Rows in
\(\mathbf{M}\) define how a given component \(i\) is affected by other
components of the system, meaning that the rate of component response
time can be modelled by multiplying all row elements by a real scalar
value
\(\gamma_{i}\)\textsuperscript{\protect\hyperlink{ref-Patel2018}{15}}.
The distribution of \(\gamma\) over \(S\) components thereby models the
distribution of component response rates. An instructive example
compares one \(\mathbf{M}\) where \(\gamma_{i} = 1\) for all \(i\) in
\(S\) to the same \(\mathbf{M}\) when half of \(\gamma_{i} = 1.95\) and
half of \(\gamma_{i} = 0.05\). This models one system in which
\(\gamma\) is invariant and one in which \(\gamma\) varies, but systems
are otherwise identical (note \(E[\gamma_{i}] = 1\) in both cases). I
assume \(S = 200\), \(C = 0.05\), and \(\sigma = 0.4\); diagonal
elements are set to \(-1\) and non-zero off-diagonal elements are drawn
randomly from \(\mathcal{N}(0, \sigma^{2})\). Rows are then multiplied
by \(\gamma_{i}\) to generate \(\mathbf{M}\). When \(\gamma_{i} = 1\),
eigenvalues of \(\mathbf{M}\) are distributed uniformly within a circle
centred at (\(-1, 0\)) with a radius of 1.265 (Fig. 1a). Hence, the real
components of eigenvalues are highly unlikely to all be negative when
all \(\gamma_{i} = 1\). But when \(\gamma_{i}\) values are separated
into two groups, eigenvalues are no longer uniformly distributed (Fig.
1b). Instead, two distinct clusters of eigenvalues appear (red circles
in Fig. 1b), one centred at (\(-1.95, 0\)) and the other centred at
(\(-0.05, 0\)). The former has a large radius, but the real components
have shifted to the left (in comparison to when \(\gamma = 1\)), while
the latter cluster has a smaller radius, but real components have
shifted to the right; all \(\Re({\lambda}) < 0\). Overall, for 1 million
randomly assembled \(\mathbf{M}\), this division between slow and fast
component response rates results in more stable systems: 1 stable given
\(\gamma = 1\) versus 32 stable given \(\gamma = \{1.95, 0.5\}\).

Higher stability in systems with variation in \(\gamma\) can be observed
by sampling \(\gamma_{i}\) values from various distributions. I focus on
a uniform distribution where \(\gamma \sim \mathcal{U}(0, 2)\) (see
Supplementary Information for other distributions, which give similar
results). As with the case of \(\gamma = \{1.95, 0.5\}\) (Fig. 1b),
\(E[\gamma] = 1\) when \(\gamma \sim \mathcal{U}(0, 2)\), allowing
comparison of \(\mathbf{M}\) before and after the addition of variation
in component response rate. Figure 2 shows a comparison of eigenvalue
distributions given \(S = 1000\), \(C = 0.05\), and \(\sigma = 0.4\). As
expected\textsuperscript{\protect\hyperlink{ref-Tao2010}{14}}, when
\(\gamma = 1\), eigenvalues are distributed uniformly in a circle
centred at (\(-1, 0\)) with a radius of \(\sigma\sqrt{SC} =\) 2.828.
Uniform variation in \(\gamma\) leads to a non-uniform distribution of
eigenvalues\textsuperscript{\protect\hyperlink{ref-Ahmadian2015}{16}--\protect\hyperlink{ref-Stone2017}{18}},
some of which are clustered locally near the centre of the distribution,
but others of which are spread outside the former radius of 2.828 (Fig
2b). The clustering and spreading of eigenvalues introduced by
\(Var(\gamma)\) can destabilise previously stable systems or stabilise
systems that are otherwise unstable. But where systems are otherwise too
complex to be stable given \(\gamma = 1\), the effect of \(Var(\gamma)\)
can often lead to stability above
May's\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}
threshold \(\sigma\sqrt{SC} < 1\).

\textbf{Simulation of random \(\mathbf{M}\) across \(\mathbf{S}\)}. To
investigate the effect of \(Var(\gamma)\) on system stability, I
simulated random \(\mathbf{M}\) matrices at \(\sigma = 0.4\) and
\(C = 1\) across \(S = \{2, 3, ..., 49, 50\}\). One million
\(\mathbf{M}\) were simulated for each \(S\), and the stability of
\(\mathbf{M}\) was assessed given \(\gamma = 1\) versus
\(\gamma \sim \mathcal{U}(0, 2)\). For all \(S > 10\), I found that the
number of stable random systems was higher given \(Var(\gamma)\) than
when \(\gamma = 1\) (Fig. 3; see Supplementary Information for full
table of results), and that the difference between the probabilities of
observing a stable system increased with an increase in \(S\); i.e., the
potential for \(Var(\gamma)\) to drive stability increased with system
complexity. For the highest values of \(S\), nearly all systems that
were stable given \(Var(\gamma)\) would not have been stable given
\(\gamma = 1\).

\textbf{Targeted manipulation of \(\mathbf{\gamma}\)}. To further
investigate the potential of \(Var(\gamma)\) to be stabilising, I used a
genetic algorithm. Genetic algorithms are heuristic tools that mimic
evolution by natural selection, and are useful when the space of
potential solutions (in this case, possible combinations of \(\gamma\)
values leading to stability in a large complex system) is too large to
search
exhaustively\textsuperscript{\protect\hyperlink{ref-Hamblin2013}{19}}.
Generations of selection on \(\gamma\) value combinations to minimise
\(\max\left(\Re(\lambda)\right)\) demonstrated the potential for
\(Var(\gamma)\) to increase system stability. Across
\(S = \{2, 3, ..., 39, 40\}\), sets of \(\gamma\) values were found that
resulted in stable systems with probabilities that were up to four
orders of magnitude higher than when \(\gamma = 1\) (Fig. 4), meaning
that stability could often be achieved by manipulating \(S\) \(\gamma\)
values rather than \(S \times S\) \(\mathbf{M}\) elements.

\textbf{System feasibility given \(\mathbf{Var(\gamma)}\)} For complex
systems in which individual system components represent the density of
some tangible quantity, it is relevant to consider the feasibility of
the system. Feasibilility assumes that the system is stable and values
of all components are positive at
equilibrium\textsuperscript{\protect\hyperlink{ref-Grilli2017}{5},\protect\hyperlink{ref-Dougoud2018}{20},\protect\hyperlink{ref-Song2018}{21}}.
This is of particular interest for ecological communities because
population density cannot take negative values, meaning that ecological
systems need to be feasible for stability to be biologically
realistic\textsuperscript{\protect\hyperlink{ref-Dougoud2018}{20}}.
While my results are intended to be general to all complex systems, and
not restricted to species networks, I have also performed a feasibility
analysis on all matrices \(\mathbf{M}\) tested for stability, and
additionally for specific types of ecological
communities\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}
(e.g., competitive, mutualist, predator-prey; see Supplementary
Information). Feasibility was unaffected by \(Var(\gamma)\), meaning
that for pure interacting species networks, variation in component
response rate (i.e., species generation time) does not affect stability
at biologically realistic species densities.

@@

\textbf{Figure 1: Example distribution of eigenvalues before (a) and
after (b) separating a randomly generated complex system into fast
(\(\boldsymbol{\gamma} = 1.95\)) and slow
(\(\boldsymbol{\gamma} = 0.05\)) component response rates.} Each panel
shows the same system where \(S = 200\), \(C = 0.05\), and
\(\sigma = 0.4\), and in each case \(E[\gamma] = 1\) (i.e., only the
distribution of \(\gamma\) differs between panels). \textbf{a.}
Eigenvalues plotted when all \(\gamma = 1\); distributions of points are
uniformly distributed within the blue circle with a radius of
\(\sigma\sqrt{SC} =\) 1.238 centred at -1 on the real axis. \textbf{b.}
Eigenvalues plotted when half \(\gamma = 1.95\) and half
\(\gamma = 0.05\); distributions of points can be partitioned into one
large circle centred at \(\gamma = -1.95\) and one small circle centred
at \(\gamma = -0.05\). In a, the maximum real eigenvalue
\(\max\left(\Re(\lambda)\right) =\) 0.2344871, while in b
\(\max\left(\Re(\lambda)\right) =\) -0.0002273135, meaning that the
complex system in b but not a is stable because in b
\(\max\left(\Re(\lambda)\right) < 0\). In 1 million randomly generated
complex systems under the same parameter values, 1 was stable when
\(\gamma = 1\) while 32 were stable when \(\gamma = \{1.95, 0.05\}\).
Overall, complex systems that are separated into fast versus slow
components tend to be more stable than otherwise identical systems with
identical component response rates.

\includegraphics{unnamed-chunk-10-1.pdf}

@@

\subsection{Discussion}\label{discussion}

Here I have shown that the stability of large systems might often be
contigent upon variation in the response rates of their individual
components, meaning that factors such as rate of trait evolution (in
biological networks), transaction speed (in economic networks), or
communication speed (in social networks) need to be considered when
investigating the stability of complex systems. Variation in component
response rate becomes more likely to be critical for stability as system
size increases, and can ultimately increase the overall probability that
system stability is observed above that predicted by
May's\textsuperscript{\protect\hyperlink{ref-May1972}{1}} classically
derived \(\sigma \sqrt{SC}\) criterion. The logic outlined here is
general, and potentially applies to any complex system in which
individual system components can vary in their reaction rates to system
perturbation.

@@

It is important to emphasise that variation in component response rate
is not stabilising per se; that is, adding variation in component
response rate to a particular system does not necessarily increase the
probability that the system will be stable. Rather, systems that are
observed to be stable are more likely to have varying component response
rates, and for this variation to be critical to their stability (Fig.
3). This is caused by the shift to a non-uniform distribution of
eigenvalues that occurs by introducing \(Var(\gamma)\) (Fig. 1b, 2b),
which can sometimes cause all of the real components of the eigenvalues
of the community matrix \(\mathbf{M}\) to become negative, but might
also increase the real components of eigenvalues. The mathematics
underlying this shift in eigenvalue distribution has been
investigated\textsuperscript{\protect\hyperlink{ref-Ahmadian2015}{16}}
and recently applied to questions concerning species density and
feasibility\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{17},\protect\hyperlink{ref-Stone2017}{18}},
but has not been interpreted as rates of response of individual system
components to perturbation.
